Multi-level generalizations of collective spin dynamics and squeezing

### 2019-09-26 sun_squeezing/sun_notes.tex
added text about an external laser drive and spin-orbit coupling

Under the frozen-mode approximation, two-body collective
SU($n$)-symmetric interactions between fermions on a lattice take the
form\cite{perlin2019effective}
\begin{align}
  H_{\t{int}} = \f{u}{2} \sum_{p,q,\mu,\nu}
  \p{c_{p\mu}^\dag c_{q\nu}^\dag c_{q\nu} c_{p\mu}
    + c_{q\mu}^\dag c_{p\nu}^\dag c_{q\nu} c_{p\mu}},
\end{align}
where $u\equiv U/L$ is the two-body on-site interaction energy $U$
divided by the number of lattice sites $L$, and $c_{p\mu}$ is the
annihilation operator for a fermionic mode with quasi-momentum $p$ and
internal state (e.g.~nuclear spin) $\mu$.  Defining the spin operators
$s_{\mu\nu}^{(p)}\equiv c_{p\mu}^\dag c_{p\nu}$ for each
quasi-momentum $p$, we can alternately write
\begin{align}
  H_{\t{int}} = \f{u}{2} \sum_{p,q,\mu,\nu}
  \p{s_{\mu\mu}^{(p)} s_{\nu\nu}^{(q)} - s_{\mu\nu}^{(p)} s_{\nu\mu}^{(q)}}
  = \f{u}{2}\p{N^2 - \sum_{p,q,\mu,\nu} s_{\mu\nu}^{(p)} s_{\nu\mu}^{(q)}},
\end{align}
where $N$ is the total number of particles.  In the absence of
coherence between sectors of different particle number, we can neglect
the $\sim N^2$ term and simply write
\begin{align}
  H_{\t{int}}
  = - \f{u}{2} \sum_{p,q,\mu,\nu} s_{\mu\nu}^{(p)} s_{\nu\mu}^{(q)}.
\end{align}
In the case of SU(2), at this point we would expand the spin operators
$s_{\mu\nu}^{(p)}$ in terms of Pauli operators in order to convert
$H_{\t{int}}$ into an SU(2) spin Hamiltonian.  To generalize this
procedure to the case of SU($n$), we first define a vector
$\v s^{(p)}$ of all spin operators $s_{\mu\nu}^{(p)}$, and write
\begin{align}
  H_{\t{int}}
  = - \f{u}{2} \sum_{p,q} {\v s^{(p)}}^\dag \v s^{(q)}
  = - \f{u}{2} \v\S^\dag \c \v\S,
  &&
  \v\S\equiv \sum_p \v s^{(p)}.
  \label{eq:H_int_spin}
\end{align}
In the following sections, we identify various bases for the space of
operators on the Hilbert space of an $n$-level system.  For brevity,
we denote this space of operators by $\B_n$, and note that $\B_n$ is
itself an $n^2$-dimensional Hilbert space equipped with a trace
(Hilbert-Schmidt) inner product
\begin{align}
  \obk{\O|\Q} \equiv \tr\p{\O^\dag Q},
\end{align}
where $\oket{\Q}$ and $\obra{\O}$ respectively denote vectors in
$\B_n$ and its dual space $\B_n^*$.  Similarly to the set of spin
operators $\set{s_{\mu\nu}}$ on the Hilbert space at each
quasi-momentum, all of the bases we consider will be orthonormal with
respect to the trace inner product.  As we show in Appendix
\ref{sec:changing_bases}, the collective SU($n$)-symmetric interaction
Hamiltonian in \eqref{eq:H_int_spin} takes an identical form in any
orthonormal basis for $\B_n$.

@@

The most obvious generalization of Pauli operators to the case of
SU($n$) are the {\it generalized Gell-Mann (GGM)
  operators}\cite{hioe1981level, bertlmann2008bloch}, defined for
$j,k,\ell\in\set{0,1,\cdots,n-1}$ with $k<j$ and $\ell>0$ by
\begin{align}
  \lambda_0 \equiv \sqrt{\f{2}{n}}~ \1,
  &&
  \lambda_\ell \equiv \sqrt{\f{2}{\ell\p{\ell+1}}}
  \p{\sum_{n=0}^{\ell-1}\op{n} - \ell \op{\ell}},
  \label{eq:GGM_diag}
\end{align}
\begin{align}
  \lambda_{jk,+} \equiv \op{j}{k} + \op{k}{j},
  &&
  \lambda_{jk,-} \equiv i\p{\op{j}{k} - \op{k}{j}},
  \label{eq:GGM_off_diag}
\end{align}
where $\1$ is the identity operator.  The self-adjoint GGM operators
in \eqref{eq:GGM_diag} and \eqref{eq:GGM_off_diag} provide a complete
basis for the space of operators on the Hilbert space of an $n$-level
system, coincide exactly with the Pauli operators in the case of
$n=2$, and satisfy the orthonormality condition
\begin{align}
  \obk{\lambda_a|\lambda_b} = 2\delta_{ab},
  \label{eq:GGM_inner}
\end{align}
where $a,b$ index any GGM operator in \eqref{eq:GGM_diag} or
\eqref{eq:GGM_off_diag}.  Denoting a vector of all GGM operators by
$\v\lambda$, we can therefore write (see Appendix
\ref{sec:changing_bases})
\begin{align}
  H_{\t{int}}
  = -\f{u}{4} \sum_{p,q} \v\lambda^{(p)} \c \v\lambda^{(q)}
  = -u \v\Lambda \c \v\Lambda,
  &&
  \v\Lambda \equiv \f12 \sum_p \v\lambda^{(p)}.
  \label{eq:H_int_GGM}
\end{align}

@@

The GGM operators in \eqref{eq:GGM_diag} and \eqref{eq:GGM_off_diag}
provide a convenient operator basis to describe dynamics obeying
SU($n$) symmetry.  An external driving field addressing $n$-level
nuclear spins, however, will generally violate SU($n$) symmetry, and
instead obeys the symmetries of the {\it polarization} or {\it
  transition operators} defined by\cite{kryszewski2006alternative,
  bertlmann2008bloch}
\begin{align}
  T_{JM} \equiv \sqrt{\f{2J+1}{2I+1}} \sum_{\mu,\nu}
  \bk{I\mu;I\nu,JM} \op{\mu}{\nu},
  \label{eq:trans_ops}
\end{align}
where $J\in\set{0,1,\cdots,n-1}$ and $M\in\set{-J,-J+1,J}$ index a
total spin and its projection onto a quantization axis;
$I\equiv\p{n-1}/2$ is the maximal angular momentum of a spin-$n$
system; $\mu,\nu\in\set{-I,-I+1,\cdots,I}$ index projections of an
$n$-level nuclear spin onto a quantization axis; and
$\bk{I\mu;I\nu,JM}$ is a Clebsch-Gordan coefficient.  Similarly to the
GGM operators in \eqref{eq:GGM_diag} and \eqref{eq:GGM_off_diag}, the
transition operators in \eqref{eq:trans_ops} provide a complete basis
for the space of operators on the Hilbert space of an $n$-level
system, and satisfy the orthonormality condition
\begin{align}
  \obk{T_{JM}|T_{J'M'}} = \delta_{JJ'} \delta_{MM'},
\end{align}
which implies
\begin{align}
  H_{\t{int}} = -\f{u}{2} \sum_{p,q} {\v T^{(p)}}^\dag \c \v T^{(q)}
  = -\f{u}{2} \v\T^\dag \c \v\T,
  &&
  \v\T \equiv \v\T\equiv\sum_p\v T^{(p)},
  \label{eq:H_int_trans}
\end{align}
where $\v T$ is a vector of all transition operators in
\eqref{eq:trans_ops}.

@@

Unlike the GGM operators in \eqref{eq:GGM_diag} and
\eqref{eq:GGM_off_diag}, the transition operators in
\eqref{eq:trans_ops} are not self-adjoint.  In order to express
Hamiltonians in terms of self-adjoint operators, we define the {\it
  drive operators}
\begin{align}
  D_{J,0} \equiv \sqrt{2}~ T_{J,0},
  &&
  D_{J\Delta,+} \equiv -\p{T_{J\Delta} + T_{J\Delta}^\dag},
  &&
  D_{J\Delta,-} \equiv i\p{T_{J\Delta} - T_{J\Delta}^\dag},
  \label{eq:drive_ops}
\end{align}
for $\Delta\in\set{1,2,\cdots,J}$.  Similarly to the GGM operators in
\eqref{eq:GGM_diag} and \eqref{eq:GGM_off_diag}, the drive operators
are equal to the Pauli operators in the case of $n=2$, and satisfy the
orthonormality condition
\begin{align}
  \obk{D_a|D_b} = 2\delta_{ab},
\end{align}
where $a,b$ index any drive operator in \eqref{eq:drive_ops}.  These
operators have the convenient feature that $D_{1,1,\pm}$ is (up to a
gauge transformation) proportional to the Hamiltonian induced on a
spin-$n$ system by a classical driving laser.  In terms of the drive
operators,
\begin{align}
  H_{\t{int}} = -\f{u}{4} \sum_{p,q} \v D^{(p)}\c\v D^{(q)}
  = -u \v\D \c \v\D,
  &&
  \v\D \equiv \f12 \sum_p \v D^{(p)},
  \label{eq:H_int_drive}
\end{align}
where $\v D$ is a vector of all drive operators in
\eqref{eq:drive_ops}.

We note that Weyl operators\cite{bertlmann2008bloch} and self-adjoint
linear combinations thereof\cite{asadian2016heisenbergweyl} also
provide an interesting basis of operators, generalizing the notion of
position and momentum to the case of a finite-dimensional Hilbert
space.  As our present work is motivated by multi-level systems that
are realized with nuclear spin (angular momentum) states, however, the
transition and drive operators are more natural to consider than the
Weyl operators, so we will not discuss the latter any further.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{External drive}

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{Spin-orbit coupling}

@@

>>>>

We consider a collection of $n$-level fermions on a 1-D lattice, with
a single-particle Hamiltonian
\begin{align}
  H_{\t{lat}} = -J \sum_{q,\mu} \cos\p{qa} c_{q\mu}^\dag c_{q\mu},
  \label{eq:H_lat}
\end{align}
where $J$ is the nearest-neighbor tunneling rate, $a$ is the spacing
between neighboring lattice sites, and $c_{q\mu}$ is the fermionic
annihilation operator for a particle in quasi-momentum mode $q$ with
internal state (e.g.~nuclear spin) $\mu$.  Under the frozen-mode
approximation, two-body collective SU($n$)-symmetric interactions
between these fermions take form\cite{perlin2019effective}
\begin{align}
  H_{\t{int}} = \f12 u \sum_{p,q,\mu,\nu}
  \p{c_{p\mu}^\dag c_{q\nu}^\dag c_{q\nu} c_{p\mu}
    + c_{q\mu}^\dag c_{p\nu}^\dag c_{q\nu} c_{p\mu}},
\end{align}
where $u\equiv U/L$ is the two-body on-site interaction energy $U$
divided by the number of lattice sites $L$.  Defining the spin
operators $s_{\mu\nu}^{(p)}\equiv c_{p\mu}^\dag c_{p\nu}$ for each
quasi-momentum $p$, we can alternately write
\begin{align}
  H_{\t{int}} = \f12 u \sum_{p,q,\mu,\nu}
  \p{s_{\mu\mu}^{(p)} s_{\nu\nu}^{(q)} - s_{\mu\nu}^{(p)} s_{\nu\mu}^{(q)}}
  = \f12 u\p{N^2 - \sum_{p,q,\mu,\nu} s_{\mu\nu}^{(p)} s_{\nu\mu}^{(q)}},
\end{align}
where $N$ is the total number of particles.  In the absence of
coherence between sectors of different particle number, we can neglect
the $\sim N^2$ term and simply write
\begin{align}
  H_{\t{int}}
  = - \f12 u \sum_{p,q,\mu,\nu} s_{\mu\nu}^{(p)} s_{\nu\mu}^{(q)}.
\end{align}
In the case of SU(2), at this point we would expand the spin operators
$s_{\mu\nu}^{(p)}$ in terms of Pauli operators in order to convert
$H_{\t{int}}$ into an SU(2) spin Hamiltonian.  To generalize this
procedure to the case of SU($n$), we first define a vector
$\v s^{(p)}$ of all spin operators $s_{\mu\nu}^{(p)}$, and write
\begin{align}
  H_{\t{int}}
  = - \f12 u \sum_{p,q} {\v s^{(p)}}^\dag \v s^{(q)}
  = - \f12 u \v\S^\dag \c \v\S,
  &&
  \v\S\equiv \sum_p \v s^{(p)}.
  \label{eq:H_int_spin}
\end{align}
In the following sections, we identify various bases for the space of
operators on the Hilbert space of an $n$-level system.  For brevity,
we denote this space of operators by $\B_n$, and note that $\B_n$ is
itself an $n^2$-dimensional Hilbert space equipped with a trace
(Hilbert-Schmidt) inner product
\begin{align}
  \obk{\O|\Q} \equiv \tr\p{\O^\dag Q},
\end{align}
where $\oket{\Q}$ and $\obra{\O}$ respectively denote vectors in
$\B_n$ and its dual space $\B_n^*$.  Similarly to the set of all spin
operators $s_{\mu\nu}\equiv\op{\mu}{\nu}$ on the single-particle
Hilbert space at each quasi-momentum, all of the bases we consider
will be orthonormal with respect to the trace inner product.  As we
show in Appendix \ref{sec:changing_bases}, the collective
SU($n$)-symmetric interaction Hamiltonian in \eqref{eq:H_int_spin}
takes an identical form in any orthonormal basis for $\B_n$.

@@

The most obvious generalization of Pauli operators to the case of
SU($n$) are the {\it generalized Gell-Mann (GGM)
  operators}\cite{hioe1981level, bertlmann2008bloch}, defined for
$j,k,\ell\in\set{0,1,\cdots,n-1}$ with $k<j$ and $\ell>0$ by
\begin{align}
  \lambda_0 \equiv \sqrt{\f{2}{n}}~ \1,
  &&
  \lambda_\ell \equiv \sqrt{\f{2}{\ell\p{\ell+1}}}
  \p{\sum_{n=0}^{\ell-1}\op{n} - \ell \op{\ell}},
  \label{eq:GGM_diag}
\end{align}
\begin{align}
  \lambda_{jk,+} \equiv \op{j}{k} + \op{k}{j},
  &&
  \lambda_{jk,-} \equiv i\p{\op{j}{k} - \op{k}{j}},
  \label{eq:GGM_off_diag}
\end{align}
where $\1$ is the identity operator.  The self-adjoint GGM operators
in \eqref{eq:GGM_diag} and \eqref{eq:GGM_off_diag} provide a complete
basis for the space of operators on the Hilbert space of an $n$-level
system, coincide exactly with the Pauli operators in the case of
$n=2$, and satisfy the orthonormality condition
\begin{align}
  \obk{\lambda_a|\lambda_b} = 2\delta_{ab},
  \label{eq:GGM_inner}
\end{align}
where $a,b$ index any GGM operator in \eqref{eq:GGM_diag} or
\eqref{eq:GGM_off_diag}.  Denoting a vector of all GGM operators by
$\v\lambda$, we can therefore write (see Appendix
\ref{sec:changing_bases})
\begin{align}
  H_{\t{int}}
  = -\f14 u \sum_{p,q} \v\lambda^{(p)} \c \v\lambda^{(q)}
  = -u \v\Lambda \c \v\Lambda,
  &&
  \v\Lambda \equiv \f12 \sum_p \v\lambda^{(p)}.
  \label{eq:H_int_GGM}
\end{align}

@@

The GGM operators in \eqref{eq:GGM_diag} and \eqref{eq:GGM_off_diag}
provide a convenient operator basis to describe dynamics obeying
SU($n$) symmetry.  An external driving field addressing $n$-level
nuclear spins, however, will generally violate SU($n$) symmetry, and
instead obeys the symmetries of the {\it polarization} or {\it
  transition operators} defined by\cite{kryszewski2006alternative,
  bertlmann2008bloch}
\begin{align}
  T_{JM} \equiv \sqrt{\f{2J+1}{2I+1}} \sum_{\mu,\nu}
  \bk{I\mu;I\nu,JM} \op{\mu}{\nu},
  \label{eq:trans_ops}
\end{align}
where $J\in\set{0,1,\cdots,n-1}$ and $M\in\set{-J,-J+1,J}$ index a
total spin and its projection onto a quantization axis;
$I\equiv\p{n-1}/2$ is the maximal angular momentum of an $n$-level
spin; $\mu,\nu\in\set{-I,-I+1,\cdots,I}$ index projections of an
$n$-level nuclear spin onto a quantization axis; and
$\bk{I\mu;I\nu,JM}$ is a Clebsch-Gordan coefficient.  Similarly to the
GGM operators in \eqref{eq:GGM_diag} and \eqref{eq:GGM_off_diag}, the
transition operators in \eqref{eq:trans_ops} provide a complete basis
for the space of operators on the Hilbert space of an $n$-level
system, and satisfy the orthonormality condition
\begin{align}
  \obk{T_{JM}|T_{J'M'}} = \delta_{JJ'} \delta_{MM'},
\end{align}
which implies
\begin{align}
  H_{\t{int}} = -\f12 u \sum_{p,q} {\v T^{(p)}}^\dag \c \v T^{(q)}
  = -\f12 u \v\T^\dag \c \v\T,
  &&
  \v\T \equiv \v\T\equiv\sum_p\v T^{(p)},
  \label{eq:H_int_trans}
\end{align}
where $\v T$ is a vector of all transition operators in
\eqref{eq:trans_ops}.

@@

Unlike the GGM operators in \eqref{eq:GGM_diag} and
\eqref{eq:GGM_off_diag}, the transition operators in
\eqref{eq:trans_ops} are not self-adjoint.  In order to express
Hamiltonians in terms of self-adjoint operators, we define the {\it
  drive operators}
\begin{align}
  D_{J,0} \equiv \sqrt{2}~ T_{J,0},
  &&
  D_{J\Delta,+} \equiv -\p{T_{J\Delta} + T_{J\Delta}^\dag},
  &&
  D_{J\Delta,-} \equiv i\p{T_{J\Delta} - T_{J\Delta}^\dag},
  \label{eq:drive_ops}
\end{align}
for $\Delta\in\set{1,2,\cdots,J}$.  Similarly to the GGM operators in
\eqref{eq:GGM_diag} and \eqref{eq:GGM_off_diag}, the drive operators
are equal to the Pauli operators in the case of $n=2$, and satisfy the
orthonormality condition
\begin{align}
  \obk{D_a|D_b} = 2\delta_{ab},
\end{align}
where $a,b$ index any drive operator in \eqref{eq:drive_ops}.  These
operators have the convenient feature that $D_{1,1,\pm}$ is (up to a
gauge transformation) proportional to the Hamiltonian induced on an
$n$-level spin by a classical driving laser.  In terms of the drive
operators,
\begin{align}
  H_{\t{int}} = -\f14 u \sum_{p,q} \v D^{(p)}\c\v D^{(q)}
  = -u \v\D \c \v\D,
  &&
  \v\D \equiv \f12 \sum_p \v D^{(p)},
  \label{eq:H_int_drive}
\end{align}
where $\v D$ is a vector of all drive operators in
\eqref{eq:drive_ops}.

We note that Weyl operators\cite{bertlmann2008bloch} and self-adjoint
linear combinations thereof\cite{asadian2016heisenbergweyl} also
provide an interesting basis of operators, generalizing the notion of
position and momentum to the case of a finite-dimensional Hilbert
space.  As our present work is motivated by multi-level systems that
are realized with nuclear spin (angular momentum) degrees of freedom,
however, the transition and drive operators are more natural to
consider than the Weyl operators, so we will not discuss the latter
any further.

%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%%
\section{External drive and spin-orbit coupling}

An external driving laser addressing a spin-$1/2$ particle on a
lattice will induce the Hamiltonian
\begin{align}
  \left. H_{\t{drive}}^{(\phi)} \right|_{n=2}
  = \f12 \Omega \sum_j e^{i\phi_j} c_{j,\up}^\dag c_{j,\dn} + \t{h.c.},
  \label{eq:H_drive_2}
\end{align}
where $\Omega$ is a drive strength, $j$ indexes a lattice site, and
$\phi_j$ is a site-dependent phase.  The multi-level analogue of
\eqref{eq:H_drive_2} is
\begin{align}
  H_{\t{drive}}^{(\phi)}
  = -\f12 \Omega \sum_j e^{i\phi_j} T_{1,1}^{(j)} + \t{h.c.}.
  \label{eq:H_drive}
\end{align}
We can make this drive spatially homogenous via the gauge
transformation $c_{j\mu} \to e^{i\mu\phi_j} c_{j\mu}$, after which the
drive becomes
\begin{align}
  H_{\t{drive}}
  = -\f12 \Omega \sum_j T_{1,1}^{(j)} + \t{h.c.}
  = \f12 \Omega \sum_j D_{1,1,+}
  = \Omega \D_{1,1,+},
\end{align}
where $\D_a\equiv\sum_jD_a^{(j)}/2$ is a collective drive operator,
with e.g.~$\D_{1,1,+}=S_\x$ in the case of $n=2$.  The SU($n$)
symmetry of collective interactions means that the interaction
Hamiltonian $H_{\t{int}}$ in \eqref{eq:H_int_drive} remains unchanged
under this gauge transformation, but the single-particle Hamiltonian
in \eqref{eq:H_lat} transformed as
\begin{align}
  H_{\t{lat}}
  \to H_{\t{lat}}^{(\phi)}
  = -J \sum_{q,\mu} \cos\p{qa+\mu\phi} c_{q\mu}^\dag c_{q\mu}
  = -J \sum_{q,\mu} \cos\p{qa+\mu\phi} s_{\mu\mu}^{(q)},
\end{align}
where $\phi$ is the relative phase between adjacent lattice sites.  In
terms of the drive operators, we can expand
\begin{align}
  H_{\t{lat}}^{(\phi)}
  = -\sum_q \sp{\cos\p{qa} \sum_{J\,\t{even}} B_J D_{J,0}^{(q)}
    - \sin\p{qa} \sum_{J\,\t{odd}} B_J D_{J,0}^{(q)}},
\end{align}
where the effective ``fields'' $B_J$ take the form
\begin{align}
  \left. B_J \right|_{J\,\t{even}}
  = \f12 \sum_\mu \obk{D_{J,0}|s_{\mu\mu}} \cos\p{\mu\phi},
  &&
  \left. B_J \right|_{J\,\t{odd}}
  = \f12 \sum_\mu \obk{D_{J,0}|s_{\mu\mu}} \sin\p{\mu\phi}.
\end{align}

@@

[not yet written]
